---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%",
  message = FALSE,
  warning = FALSE
)
```

# distributions

<!-- badges: start -->
[![Travis build status](https://travis-ci.org/alexpghayes/distributions.svg?branch=master)](https://travis-ci.org/alexpghayes/distributions)
<!-- badges: end -->

`distributions` is inspired by the [eponynmous Julia package](https://github.com/JuliaStats/Distributions.jl), which takes a generic function approach to probability distributions.

The key idea of `distributions` is to replace the `rnorm()`, `pnorm()`, etc, family of functions with S3 methods for distribution objects. The main generics are:

- `random()`: Draw samples from a distribution.
- `pdf()`: Evaluate the probability density of mass at a point `x`.
- `cdf()`: Evaluate the cumulative probability up to a point `x`.
- `quantile()`: Determine the quantile at `p`. Inverse of `cdf()`.

Another central goal of `distributions` is to use more intuitive and explicit names, to make it easier for students in intro stats classes to manipulate probabilities.

One consequence of this goal is that you can find extensive documentation that might normally belong in a vignette the documentation of functions themselves. For example:

```{r}
?normal
```

This documentation provides information on the normal distribution, and includes worked examples of how to perform Z-tests and calculate confidence intervals. We've been carefully to make sure that the math renders in a readable manner both in RStudio's Help pane, and also on the `pkgdown` website.

## Installation

`distributions` is not yet on CRAN. You can install the development version with:

``` r
install.packages("devtools")
devtools::install_github("alexpghayes/distributions")
```

## Example

The basic usage of `distributions` looks like:

```{r intro}
library(distributions)

b <- bernoulli(0.1)

random(b, 10)
pdf(b, 1)
cdf(b, 0)
quantile(b, 0.5)
```

Note that `quantile()` always returns lower tail probabilities.

## Development roadmap

The initial goal of `distributions` is to replace the base R `d`, `p`, `q`, and `r` functions for the following families:

- [x] `beta` - documented
- [x] `binom` - documented
- [x] `cauchy`
- [x] `chisq`
- [x] `exp`
- [x] `f`
- [x] `gamma`
- [x] `norm`
- [x] `pois` - documented
- [ ] `t`
- [ ] `multinomial`
- [ ] `unif`
- [ ] `geom`
- [ ] `hyper`
- [ ] `logis`
- [ ] `lnorm`
- [ ] `nbinom`
- [ ] `weibull`
- [ ] `tukey`

After that the goal will probably to support more generics like those in the `Distributions.jl`, with the hope that the community will contribute additional distributions.

Candidates for the next set of generics:

- `log_pdf()`
- `log_cdf()`
- `plot_pdf()`
- `plot_cdf()`

## Contributing

The best way to get started is to look at `R/beta.R` and `tests/testthat/test-beta.R`, copy them, and modify them for whatever new distribution you'd like to add.

Please note that `distributions` is released with a
[Contributor Code of Conduct](.github/CODE_OF_CONDUCT.md).
By contributing to this project, you agree to abide by its terms.

## Related work

For a comprehensive overview of the many packages providing various distribution related functionality see the [CRAN Task View](https://cran.r-project.org/web/views/Distributions.html).

- [`distr`](http://distr.r-forge.r-project.org/) is quite similar to `distributions`, but uses S4 objects
- [`fitdistrplus`](https://cran.r-project.org/web/packages/fitdistrplus/index.html) provides extensive functionality for fitting various distributions but does not treat distributions themselves as objects
