---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%",
  message = FALSE,
  warning = FALSE
)
```

# distributions

<!-- badges: start -->
[![Travis build status](https://travis-ci.org/alexpghayes/distributions.svg?branch=master)](https://travis-ci.org/alexpghayes/distributions)
<!-- badges: end -->

`distributions`, inspired by the [eponynmous Julia package](https://github.com/JuliaStats/Distributions.jl), provides a generic function interface to probability distributions.

`distributions` has two goals:

1. Replace the `rnorm()`, `pnorm()`, etc, family of functions with S3 methods for distribution objects

2. Be extremely well documented and friendly toward students in intro stat classes.

The main generics are:

- `random()`: Draw samples from a distribution.
- `pdf()`: Evaluate the probability density of mass at a point `x`.
- `cdf()`: Evaluate the cumulative probability up to a point `x`.
- `quantile()`: Determine the quantile at `p`. Inverse of `cdf()`.

## Quick links for students

- `vignette("intro-to-hypothesis-testing")`
- `vignette("one-sample-z-test")`
- `vignette("one-sample-z-confidence-interval")`

## Installation

`distributions` is not yet on CRAN. You can install the development version with:

``` r
install.packages("devtools")
devtools::install_github("alexpghayes/distributions")
```

## Example

The basic usage of `distributions` looks like:

```{r intro}
library(distributions)

b <- bernoulli(0.1)

random(b, 10)
pdf(b, 1)
cdf(b, 0)
quantile(b, 0.5)
```

Note that `quantile()` **always** returns lower tail probabilities. If you aren't sure what this means, please read the last several paragraphs of `vignette("one-sample-z-confidence-interval")` and have a gander at the plot.

## Development roadmap

The initial goal of `distributions` is to replace the base R `d`, `p`, `q`, and `r` functions for the following families:

- [x] `beta` - documented
- [x] `binom` - documented
- [x] `cauchy` - documented
- [x] `chisq`
- [x] `exp`
- [x] `f`
- [x] `gamma`
- [x] `norm` - documented
- [x] `pois` - documented
- [ ] `t`
- [ ] `multinomial`
- [ ] `unif`
- [ ] `geom`
- [ ] `hyper`
- [ ] `logis`
- [ ] `lnorm`
- [ ] `nbinom`
- [ ] `weibull`
- [ ] `tukey`

Documentation -- add an explicit section on transformations and relations to
other random variables?

We also want to write detailed vignettes describing the most common use cases for each of these distributions. Realistically, this means we need to document how to perform the statistical tests most common in intro stats classes.

This documentation should live in vignette that explicitly write down both math and code in a cookbook style mashup. These cookbooks should also provide code and guidance for basic assumption checking. See `vignette("one-sample-z-test")` for an example. The vignette wishlist is currently as follows:

- [ ] sign tests using the binomial distribution
- [ ] Z test for population proportion
  - add to existing Z-test vignette
- [ ] power of one/two sample Z test
- [ ] power of one/two sample T test (Welch's)
  - also via `power.t.test()`
- [ ] one-way ANOVA using the F distribution
  - also show `aov()` and `anova(lm())` options
  - power of anova with `power.anova.test()`
- [ ] one sample T test
  - also show `t.test()`
- [ ] two sample equal variance T-test
  - also via `t.test()`
- [ ] two sample Welch's T-test
  - also `t.test()`
- [ ] paired Z, T and sign tests
- [ ] chi square test for equal variance
- [ ] asymptotic likelihood ratio using the chi squared distribution (?)
- [ ] asymptotic wald tests using the chi squared distribution (?)
- [ ] asymptotic rao score tests using the chi squared distribution (?)



After that the goal will probably to support more generics like those in the `Distributions.jl`, with the hope that the community will contribute additional distributions.

Candidates for the next set of generics:

- `log_pdf()`
- `log_cdf()`
- `plot_pdf()`
- `plot_cdf()`

## Contributing

The best way to get started is to look at `R/beta.R` and `tests/testthat/test-beta.R`, copy them, and modify them for whatever new distribution you'd like to add.

Please note that `distributions` is released with a
[Contributor Code of Conduct](.github/CODE_OF_CONDUCT.md).
By contributing to this project, you agree to abide by its terms.

## Related work

For a comprehensive overview of the many packages providing various distribution related functionality see the [CRAN Task View](https://cran.r-project.org/web/views/Distributions.html).

- [`distr`](http://distr.r-forge.r-project.org/) is quite similar to `distributions`, but uses S4 objects
- [`fitdistrplus`](https://cran.r-project.org/web/packages/fitdistrplus/index.html) provides extensive functionality for fitting various distributions but does not treat distributions themselves as objects
