---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```
# distributions

<!-- badges: start -->
<!-- badges: end -->

`distributions` is inspired by the [eponynmous Julia package](https://github.com/JuliaStats/Distributions.jl), which takes a generic function approach to probability distributions.

The key idea of `distributions` is to replace the `rnorm()`, `pnorm()`, etc, family of functions with S3 methods for distribution objects. The main generics are:

- `random()`: Draw samples from a distribution.
- `pdf()`: Evaluate the probability density of mass at a point `x`.
- `cdf()`: Evaluate the cumulative probability up to a point `x`.
- `quantile()`: Determine the quantile at `p`. Inverse of `cdf()`.

Another central goal of `distributions` is to use more intuitive and explicit names, to make it easier for students in intro stats classes to manipulate probabilities.

## Installation

`distributions` is not yet on CRAN. You can install the development version with:

``` r
install.packages("devtools")
devtools::install_github("alexpghayes/distributions")
```

## Example

The basic usage of `distributions` looks like:

```{r intro}
library(distributions)

b <- bernoulli(0.1)

random(b, 10)
pdf(b, 1)
cdf(b, 0)
quantile(b, 0.5)
```

## Development roadmap

The initial goal of `distributions` is to replace the base R `d`, `p`, `q`, and `r` functions for the following families:

- [x] `beta`
- [ ] `binom`
- [ ] `cauchy`
- [ ] `chisq`
- [ ] `exp`
- [ ] `f`
- [ ] `gamma`
- [ ] `geom`
- [ ] `hyper`
- [ ] `logis`
- [ ] `lnorm`
- [ ] `nbinom`
- [ ] `norm`
- [ ] `pois`
- [ ] `t`
- [ ] `unif`
- [ ] `weibull`
- [ ] `tukey`
- [ ] `signrank`
- [ ] `wilcox`

After that the goal will probably to support more generics like those in the `Distributions.jl`, with the hope that the community will contribute additional distributions.

Candidates for the next set of generics:

- `log_pdf()`
- `log_cdf()`

## Contributing

The best way to get started is to look at `R/beta.R` and `tests/testthat/test-beta.R`, copy them, and modify them for whatever new distribution you'd like to add.

Please note that the 'distributions' project is released with a
[Contributor Code of Conduct](.github/CODE_OF_CONDUCT.md).
By contributing to this project, you agree to abide by its terms.

## Related work

For a comprehensive overview of the many packages providing various distribution related functionality see the [CRAN Task View](https://cran.r-project.org/web/views/Distributions.html).

- [`distr`](http://distr.r-forge.r-project.org/) is quite similar to `distributions`, but uses S4 objects
- [`fitdistrplus`](https://cran.r-project.org/web/packages/fitdistrplus/index.html) provides extensive functionality for fitting various distributions but does not treat distributions themselves as objects


## Conventions

This section of the README is mostly a note to myself.

- `quantile()` always uses lower tail probabilities.
