---
title: "Paired tests"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Paired tests}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE
)
```

This vignette will make most sense if you are familiar with one sample and two sample tests. If these are new to you, have a look at `vignette("one-sample-z-test")` and `vignette("two-sample-z-test")`.

## Problem setup

Suppose we're interested in people's happiness before and after discovering the #rstats Twitter community. We survey them get before and after happiness scores, using a happiness survey that goes from 0 to 100.

The data looks like this

```{r}
set.seed(27)

before <- round(rnorm(10, 50, 20))
after <- round(before + 5 + rnorm(10, 0, 10))

knitr::kable(data.frame(person = 1:10, before, after))
```

## Working with the differences

Crucially, since the before measurement and the after measurement are on the *same person*, we can subtract the before measurement from the after measurement and do a one sample test on these differences.

```{r}
before <- c(88, 73, 35, 21, 28, 56, 50, 73, 93, 55)
after <- c(80, 78, 56, 28, 26, 50, 39, 67, 98, 63) 

diff <- after - before
diff
```



## Paired tests versus two sample tests

why pair - power
pair if you can
memorize and analogize

## Paired Z-test

## Paired T-test

## Paired sign-test

